# A Bayesian Alternative to Hierarchal Clustering 

# Authors: Thomas Yan, Mark Christopher Uy, Gregory Reid, Jose Carlos Dairo

## Abstract

A common problem in many analyses is determining how to group data, which observations belong to which group and how many groups are appropriate. In these problems, often the true grouping and group memberships of each observation are unknown and the data for that is simply unavailable. This is the reason why clustering methods are required to solve these sorts of problems, by creating groups using similarities between observations. An example of such a problem is trying to determine the number of sub-species of an animal in a habitat when the sub-species can not be easily distinguished visually. 

One method of clustering is using Bayesian inference to estimate the posterior probability to assign the most likely group for each observation as its designated cluster. A common method of estimating probabilities is Monte Carlo methods, however there are cases where its impossible to generate the independent and identically distributed samples required for Monte Carlo. 

This report's goal is to create a Markov Chain Monte Carlo (MCMC) method, specifically using a Gibbs Sampler approach, to perform clustering on data with unknown responses by estimating a posterior distribution for the class memberships. By using MCMC instead of just Monte Carlo methods, this approach is able to work in more problems and is able to estimate posterior probabilities in situations where Monte Carlo methods would not be possible. 


## Introduction 

When trying to group data where the true group memberships, if they exist, are unknown, then methods used to perform this task fall under the umbrella term, clustering, which itself is a subset of unsupervised machine learning (B. Zhang, 2003). There are already many well known methods of clustering which are widely used and effective in many contexts. These include, k-means clustering, mean-shift clustering, density-based clustering and heirarchal clustering. However, a prominent drawback for these methods is that they are heuristic and thus have no way of drawing formal inferences from their results (K. M. A. Patel and P. Thakral, 2016). 

When we elect to use model-based clustering, it is possible to draw interpretations from a statistical point of view (Bouveyron, 2013). A common approach is the maximum likelihood method where one would estimate the model parameters by maximizing the log likelihood function using an Expectation Maximization (EM) Algorithm and then estimating cluster memberships using the maximum a posterioiri rule (A.Samé, 2009).

As dimensionality and data size increases, the computational time needed to perform EM increases immensely. Neal and Hitton introduced a way to speed up the EM algorithm by recalculating the distribution for only one unobserved variable at each expectation step as opposed to calculating for all unobserved variables (R.M. Neal and Hinton, 1998). Bradley et al. proposed an implementaion of the EM Algorithm that was scalable to large databases with limited memory buffers (Bradley et al., 1998). While these solutions focused on improving the EM algorithm for specific cases, our paper is focused on developing an approach which does not involve any EM algorithm. We propose a solution that uses an exclusively bayesian approach by implementing a Gibb's sampler to estimate the parameters of our model as well as estimate the cluster memberships of our observations. The following sections will showcase what assumptions were made to make the simplify the problem enough to be able to use of a gibb's sampler, its performance on a real data set and lastly a discussion on the merits and flaws of our proposed solution.  

## Methodology 


The first assumption this method makes is that the observed data follows a mixture normal distribution, $x_i \sim \sum_{k=1}^K \rho_k N(\mu_k, \Sigma_k), i = 1, ..., N$ and are independent and identically distributed. Then following this, we take a Bayesian inference approach where the unknown, but true cluster memberships follows a multinomial distribution, $z_i \sim \text{Multinomial}(K, \rho)$ where $\rho = (\rho_1, ..., \rho_K)$ and $(N_1, ..., N_k) \sim \text{Multinomial}(N, \rho)$. However, a problem with this is that it quickly suffers from the curse of dimensionality as the dimension of $x_i$ increases and causes this mixture-normal model to fail. 

The solution to this was to instead assume that the x's belonged to a family of models $f(x_i|\theta_i)$, where the distribution of each $x_i$ depended on its own parameter $\theta_i$. Then instead of assuming that the x's followed a mixture-normal, we assume that the $\theta$'s follow a mixture-normal. Where $\theta_i \in \rm{I\!R}^p$, such that p is much smaller than N and will not severely suffer from the curse of dimensionality. As in: 

$$
\theta_i \overset{iid}{\sim} \sum_{k=1}^K \rho_k N(\mu_k, \Sigma_k), i = 1, ..., N  
$$
$$
x_i | \theta_i \overset{ind}{\sim} f(x_i | \theta_i) 
$$

Where $\theta_i$ and $f(x_i|\theta_i)$ are chosen based on the data and problem at hand and needs to be decided before the Gibbs sampler can be used. In this setup, $\theta_i$ is an unknown and random parameter and this assumed model is also known as a "random-effects" model. The standard fixed and unknown parameters that are used to define cluster sizes and membership probabilities are $\psi_k = (\rho_k, \mu_k, \Sigma_k)$.  

However, the problem is still too difficult to deal with, and impossible to do analytically, and requires further simplifcation. Let $\hat{\theta}_i = \text{arg max}_{\theta} f(x_i|\theta)$, the maximum likelihood estimate of $\theta_i$ from our observed data, and $\hat{V}_i = - [\frac{\partial}{\partial \theta^2} \log(f(x_i | \hat{\theta}_i))]^{-1}$ denote the observed Fisher information. 

Then we can approximate our above clustering model simply using the asymptotic properties of MLE which gives the following results:

$$
\theta_i \overset{iid}{\sim} \sum_{k=1}^K \rho_k N(\mu_k, \Sigma_k), i = 1, ..., N 
$$
$$
\hat{\theta}_i | \theta \overset{ind}{\sim} N(\theta_i, \hat{V}_i)
$$

This gives the result: $\hat{\theta}_i|\{z_i = k\} \sim N(\mu_k, \hat{V}_i + \Sigma_k)$. Then using this and the previous assumptions, and letting $\hat{\theta}_i = y_i$ be our observed data, we have the result: 

$$
z_i \overset{iid}{\sim} \text{Multinomial}(K, \rho) 
$$
$$
\theta_i | z_i \overset{ind}{\sim} N(\mu_{z_i}, \Sigma_{z_i}) 
$$
$$
y_i | \theta_i \overset{ind}{\sim} N(\theta_i, V_i) 
$$

Where $Y = (y_1, ..., y_N) = (\hat{\theta}_1, ... ,\hat{\theta}_N)$ is the observed data, and both $z = (z_1, ..., z_N)$ and $\Theta = (\theta_1, ..., \theta_N)$ are the missing, unobserved data. Then we can construct a complete likelihood function for our parameters $\Psi = (\psi_1, ..., \psi_K), \psi_k = (\rho_k, \mu_k, \Sigma_k)$, by assuming that the missing data was observed: 

$$
l(\Psi|z, \Theta, Y) = \log [p(Y, \Theta, z,| \Psi)] = \log[\Pi_{i=1}^N p(y_i| \theta_i, \Psi) * p(\theta_i|z_i, \Psi) * p(z_i|\Psi)]
$$

Where we know the distributions of each of these density functions from our previous assumptions. Now, the problem is to estimate the parameters when we have missing data, which will be done using Bayesian inference and an noninformative improper prior: 

$$
\pi(\Psi) ~ \alpha ~ \Pi_{k=1}^K |\Sigma_k|^{-(\nu_k + p + 1) /2} \exp(-\frac{1}{2}\text{tr}(\Sigma_k^{-1} \Omega_k))
$$
Where p is the size of each $\theta_i$, $\Omega_k$ is a p by p symmetric positive definite matrix and $\nu_k$ are constants. There are certain ways to choose $\Omega_k$ and $\nu_k$, but we will skip the details.  

Then using the prior and the above log likelihood, we can construct a posterior probability and a respective likelihood, $p(\Psi, z, \Theta|Y)$ $\alpha$ $p(Y, \Theta, z|\Psi) \pi(\Psi)$, using Bayes. Then we can simplify this posterior likelihood by conditioning on different parameters. 

If we let $\Phi = (Y, \Theta, z, \Psi)$ denote all the variables in the model and $\Phi \backslash \{\theta_i\}$ denote all variables excluding $\theta_i$. Then by simplifying the complete posterior likelihood by assuming all given terms are constant, it can be shown that:

$$
\theta_i | \Phi \backslash \{\theta_i\} \sim N(G_i(y_i - \mu_{z_i}) + \mu_{z_i}, G_i V_i) \text{ where } G_i = \Sigma_{z_i} (V_i + \Sigma_{z_i})^{-1}
$$

And that when conditioned on all other variables, all $\theta_i$'s are conditionally independent, as in $\theta_i|\Phi \backslash \{\theta_i\} \overset{\text{distribution}}{=} \theta_i|\Phi \backslash \{\Theta\}$. 

A similar result can be shown for the other variables where: 

$$
\mu_k | \Phi \backslash \{\mu_k\} \overset{ind}{\sim} N(\bar{\theta}_k, \Sigma_k / N_k) \text{ where } N_k = \sum_{i=1}^N \rm I\!I (z_i = k) 
$$
$$
\Sigma_k | \Phi \backslash \{\Sigma_k\} \overset{ind}{\sim} \text{InvWish}(\Omega_k + \sum_{i:z_i=k} (\theta_i - \mu_K)(\theta_i - \mu_k)', N_k + \nu_k) 
$$
$$
\rho_k | \Phi \backslash \{\rho_k\} \overset{ind}{\sim} \text{Dirichlet}(\alpha) \text{ where } \alpha = (N_1 + 1, ..., N_K + 1) 
$$
$$
z_i | \Phi \backslash \{z_i\} \overset{ind}{\sim} \text{Multinomial}(K, \lambda_i) \text{ where } \lambda_{ik} = \frac{\exp(\kappa_{ik})}{\sum_{m=1}^K \exp(\kappa_{im})} 
$$

Where $\kappa_{ik} = \log(P(z_i = k|\Phi \backslash \{z_i\}))= \log(\rho_k) - \frac{1}{2}[(\theta_i - \mu_k)'\Sigma_k^{-1}(\theta_i - \mu_k)' + \log |\Sigma_k|]$ 

And each of these variables are conditionally independent, which allows us to sample everything in 5 steps: simultaneously sample all $\theta_i, i = 1,..., n$, then simultaneously sample all $mu_k, k = 1, ..., K$, and so on and so forth for each of the variables. Where we repeat these 5 steps every iteration of the Gibbs sampler, for M desired iterations. 

Then we can estimate the posterior probability by $P(z_i = k|Y) \approx \hat{\lambda}_{ik}$ where $\hat{\lambda}_{ik} = [\hat{\Lambda}]_{ik}, \hat{\Lambda} = \frac{1}{M} \sum_{m=1}^M \Lambda^{(m)}$. Where $\Lambda^{(m)}$ is the $\Lambda$ matrix created in the mth iteration used to sample z. 

## Results 

```{r echo=FALSE}
init_z <- function(y, K, max_iter = 10, nstart = 10) {
  # init++
  N <- nrow(y)
  p <- ncol(y)
  x <- t(y) # easier to use columns
  centers <- matrix(NA, p, K) # centers
  icenters <- rep(NA, K-1) # indices of centers
  minD <- rep(Inf, N) # current minimum distance vector
  # initialize
  icenters[1] <- sample(N,1)
  centers[,1] <- x[,icenters[1]]
  for(ii in 1:(K-1)) {
    D <- colSums((x - centers[,ii])^2) # new distance
    minD <- pmin(minD, D)
    icenters[ii+1] <- sample(N, 1, prob = minD)
    centers[,ii+1] <- x[,icenters[ii+1]]
  }
  centers <- t(centers)
  colnames(centers) <- rownames(x)
  # run kmeans with these centers
  km <- kmeans(x = y, centers = centers, nstart = nstart, iter.max = max_iter)
  km$cluster
}
```

```{r echo=FALSE}
set.seed(440)
source("gibbs-wrapper.R")
require("plot3D")
# install.packages("plot3D") # sometimes require() does not install the plot3D package, if so, run this line
```

To demonstrate that the gibbs sampler can detect distinguishable clusters, we simulated numerical data such that the clusters are clear separated from each other.


```{r echo=FALSE}
#--------------------------------------------
# generate simulated data
# -------------------------------------------
n <- 600
A <- runif(n = 200,min = -1, max = 1)
B <- runif(n = 200, min = 100, max = 110)
C <- runif(n = 200, min = -200, max = -190 )
D <- runif(n = n, min = -5, max = 5)
# note that the data generated and stored in variables A,B,C are clearly in 3 distinct clusters
sim_data <- matrix(nrow = n,ncol = 2)
sim_data[,1] <- c(A,B,C)
sim_data[,2] <- D
var_sim1 <- var(sim_data[,1])
var_sim2 <- var(sim_data[,2])
cov_sim <- cov(sim_data[,1],sim_data[,2])
# inverse of fisher obvs is the asymptotic covariance matrix,
# thus  we will use the inverse of the observed covariance matrix as an estimate for the fisher obvs
sim_obs_info <- solve(matrix(data = c(var_sim1,cov_sim,
                                cov_sim,var_sim2),nrow = 2,ncol=2,byrow=TRUE))
plot(x=sim_data[,1],y=sim_data[,2], main = "simulated data")
```

Now we use the proposed Gibbs Sampler to determine the three clusters in this dataset.

```{r echo=FALSE, results="hide"}
# fit gibbs sampler
gibbs_fit2 <- gibbsSampler(data=sim_data, V=sim_obs_info, burnin_period=1e2, numIter=1e3, K=3, Theta.out=FALSE, z.out=FALSE)
# get estimated cluster affiliation from gibbs sampler
fitted_clusters4 <- numeric(length=n)
for (i in 1:600){
  fitted_clusters4[i] <- which.max(gibbs_fit2$Lambda[i, ])
}
# indices
index1 <- 1:n
# error rate
sim_results <- as.data.frame(sim_data)
sim_results$col <- fitted_clusters4
sim_results$col[sim_results$col == 1] <- "red"
sim_results$col[sim_results$col == 2] <- "green"
sim_results$col[sim_results$col == 3] <- "blue"
plot(x=sim_results$V1,y = sim_results$V2,col = sim_results$col, main="clusters determined by gibbs sampler")
```

In reality, it is often difficult to determine if there are distinct clusters in the dataset. If the data has many dimensions, it is very difficult to visually determine if there are distinct clusters, as was the case with the data simulated earlier. 

A real dataset that we will use in this case is dataset used by researchers trying to model microparticle trajectories with a linear drift fractional Brownian motion model (Hill et al., 2016).

This dataset contains the parameter estimates to the trajectories of 668 microparticles, their weight percentage of human bronchial epithelial (hbe) mucosal fluid, and their observed fisher information matrices. 

In this particular dataset, all observations had weight percentage of hbe as 1.5%, 2%, 2.5%, 3%, 4%, 5%.

We will see the gibbs sampler can the group the observations into 6 clusters and see if each cluster corresponds to a particular weight percentage of hbe.

```{r echo=FALSE}
hbe_fit <- readRDS("hbe-fbm_fit_v1.rds")
N <- length(hbe_fit)
p <- length(hbe_fit[[1]]$coef)
# Extract data 
wts <- vector(length=N)
theta_hat <- matrix(nrow=N, ncol=p)
colnames(theta_hat) <- names(hbe_fit[[1]]$coef) 
obs_info <- array(dim=c(p, p, N))
for (i in 1:N){
  theta_hat[i, ] <- hbe_fit[[i]]$coef
  wts[i] <- hbe_fit[[i]]$wt
  obs_info[, , i] <- hbe_fit[[i]]$vcov
}
# Transform the wts into numeric labels
lab <- unique(wts) 
K <- length(lab)
cluster <- numeric(length=N)
for (i in 1:N){
  if (wts[i] == lab[1]){
    cluster[i] <- 1
  }
  else if (wts[i] == lab[2]){
    cluster[i] <- 2
  }
  else if (wts[i] == lab[3]){
    cluster[i] <- 3
  }
  else if (wts[i] == lab[4]){
    cluster[i] <- 4
  }
  else if (wts[i] == lab[5]){
    cluster[i] <- 5
  }
  else{
    cluster[i] <- 6
  }
}
# getting true cluster affiliations
real_N_k <- numeric(length=K)
real_cols <- rep(NA,N) # colors for true cluster affiliations
for (k in 1:K){
  real_N_k[k] <- sum(cluster == k)
}
# assigning colours to each cluster
real_cols[cluster == 1] = "red"
real_cols[cluster == 2] = "blue"
real_cols[cluster == 3] = "green"
real_cols[cluster == 4] = "orange"
real_cols[cluster == 5] = "black"
real_cols[cluster == 6] = "pink"
pairs(theta_hat,col=real_cols, main="pairs plot") # pairs plot
```

Based on the pairs plot of the data, coloured based on the true weight clustering. It looks as though there is little promise to clustering based on the provided data. For mu1, mu2, lambda, and nu, the pairs plots of each against weight shows that they all overlap, that there is little promise to clustering using these variables. The within cluster ranges for the predictor values seems to heavily overlap across weight clusters which will likely make it very difficult to do any proper clustering. 

For kappa, there looks to be potential to do some basic clustering, where kappa seems to have some potential to split the data into cluster 6 vs not cluster 6. Similarly, logD seems like it could potentially make clusters of cluster 1, cluster 6 or the rest. 

However, these possibilities are assuming that the model knows the true cluster memberships. The Gibbs sampler we created is for a clustering method, which means that the true cluster label is unknown. Thus, ignoring the first column and first row of the pairs plot and removing the colouring indicating the true clusters, we can look at a new pairs plot. 


```{r, fig.width=7, fig.height=5, echo=FALSE}
pairs(theta_hat)
```

This pairs plot is much closer to the actual clustering problem where none of the true clustering behaviour is known. In all of the pairs plots, there does not seem to be any unique pairwise relationship between the predictors. Many of the pairwise plots simply look like round black blobs which will likely make it difficult to achieve a good classification accuracy. 

Despite this, we will see what clusters are determined by the gibbs sampler, and if they match the clusters determined by hbe weight percentage.


```{r results="hide"}
# fit gibbs sampler
gibbs_fit <- gibbsSampler(data=theta_hat, V=obs_info, burnin_period=1e3, numIter=1e4, K=6, Theta.out=FALSE, z.out=FALSE)
```

```{r echo=FALSE}
# get estimated cluster affiliation from gibbs sampler
fitted_clusters <- numeric(length=N)
for (i in 1:N){
  fitted_clusters[i] <- which.max(gibbs_fit$Lambda[i, ])
}
# error rate
err <-  sum(cluster != fitted_clusters)/ N
print(paste0("Percentage of estimated cluster memberships that do not match true cluster memberships: ", round(err,5)))
```

With an error rate of over 95% it is clear that the gibbs sampler did not pick out the same clusters as those determined by the weight percentage. 

As a comparison, we will see if the kmeans plus plus (km++) algorithm performs better. 

As a brief overview of this algorithm is as follows:

Step 1: Choose one center uniformly at random from among the data points.

Step 2: For each data point $x$, $d(x)$, the distance between x and the nearest center that has already been chosen.

Step 3: Choose one new data point at random as a new center, using a weighted probability distribution where a point $x$ is chosen with probability proportional to $d(x)^2$.


Step 4: Repeat Steps 2 and 3 until $k$ centers have been chosen.

Step 5: With the initial centers have been chosen, proceed using the regular k-means clustering.

(Arthur D., 2007)

```{r echo=FALSE}
# clustering real data with kmeans++ algorithm
# note since kmeans starts with random centroids, individual kmeans can have different results.
# We will conduct many kmeans and aggregate the results
#' Initial cluster allocation.
#'
#' Initializes the clusters using the kmeans++ algorithm of Arthur & Vassilvitskii (2007).
#'
#' @param y An `N x p` matrix of observations, each of which is a column.
#' @param K Number of clusters (positive integer).
#' @param max_iter The maximum number of steps in the [stats::kmeans()] algorithm.
#' @param nstart The number of random starts in the [stats::kmeans()] algorithm.
#'
#' @return A vector of length `N` consisting of integers between `1` and `K` specifying an initial clustering of the observations.
#' @author Martin Lysy
#' @references Arthur, D., Vassilvitskii, S. "k-means++: the advantages of careful seeding" *Proceedings of the 18th Annual ACM-SIAM Symposium on Discrete Algorithms.* (2007): 1027–1035. <http://ilpubs.stanford.edu:8090/778/1/2006-13.pdf>.
init_z <- function(y, K, max_iter = 10, nstart = 10) {
  # init++
  N <- nrow(y)
  p <- ncol(y)
  x <- t(y) # easier to use columns
  centers <- matrix(NA, p, K) # centers
  icenters <- rep(NA, K-1) # indices of centers
  minD <- rep(Inf, N) # current minimum distance vector
  # initialize
  icenters[1] <- sample(N,1)
  centers[,1] <- x[,icenters[1]]
  for(ii in 1:(K-1)) {
    D <- colSums((x - centers[,ii])^2) # new distance
    minD <- pmin(minD, D)
    icenters[ii+1] <- sample(N, 1, prob = minD)
    centers[,ii+1] <- x[,icenters[ii+1]]
  }
  centers <- t(centers)
  colnames(centers) <- rownames(x)
  # run kmeans with these centers
  km <- kmeans(x = y, centers = centers, nstart = nstart, iter.max = max_iter)
  km$cluster
}
max_iters <- 10000  # set number of kmeans to run
kmeans_results <- matrix(nrow = N,ncol = max_iters) # pre-allocate matrix
# simple function to calculate the mode in vector of numbers
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
# get results from max_iters kmeans
for (iter in 1:max_iters){
  kmeans_results[,iter] <- init_z(y = theta_hat,K = 6)
}
fitted_clusters1 <- matrix(nrow = N,ncol = 1) # pre-allocate matrix to store estimated clusters
# the mode cluster membership for each observation becomes the estimate cluster membership
for (row in 1:N){
  fitted_clusters1[row,] <- Mode(kmeans_results[row,])
}
err1 <- sum(cluster != fitted_clusters1) / N
print(paste0("Percentage of estimated cluster memberships that do not match true cluster memberships: ", round(err1,5)))
```

While the km++ algorithm performed better than the gibbs sampler, it still had an error rate of 70%, which means that it also did not pick out the same clusters as those determined by the weight percentage. 

It is possible that this is caused by the data being unscaled. Thus, we carry out the same test but the data has been standardized so that mean of the data is 0 and the variance is 1.

```{r echo=FALSE, results="hide"}
#-----------------------------------------------------------------------------------------------------------
# testing with standardized data (data centered at 0 with sd of 1)
#-----------------------------------------------------------------------------------------------------------
theta_hat_scaled <- scale(theta_hat)
# fit gibbs sampler
gibbs_fit1 <- gibbsSampler(data=theta_hat_scaled, V=obs_info, burnin_period=1e3, numIter=1e4, K=6, Theta.out=FALSE, z.out=FALSE)
```

```{r echo=FALSE}
# get estimated cluster affiliation from gibbs sampler
fitted_clusters2 <- numeric(length=N)
for (i in 1:N){
  fitted_clusters2[i] <- which.max(gibbs_fit1$Lambda[i, ])
}
# error rate
err2 <- sum(cluster != fitted_clusters2) / N
print(paste0("Percentage of estimated cluster memberships that do not match true cluster memberships: ",round(err2,5)))
```

The performance of the gibbs sampler improved with the error rate decreasing to 77% but is still poor.

```{r echo=FALSE}
max_iters <- 10000
kmeans_results1 <- matrix(nrow = N,ncol = max_iters)
# get results from max_iters kmeans
for (iter in 1:max_iters){
  kmeans_results1[,iter] <- init_z(y = theta_hat_scaled,K = 6)
}
fitted_clusters3 <- matrix(nrow = N,ncol = 1) # pre-allocate matrix
# the mode cluster membership for each observation becomes the estimate cluster membership
for (row in 1:N){
  fitted_clusters3[row,] <- Mode(kmeans_results1[row,])
}
err3 <- sum(cluster != fitted_clusters3) / N
print(paste0("Percentage of estimated cluster memberships that do not match true cluster memberships: ",round(err3,5)))
```

The km++ algorithm perform even worse with the error rate increasing to 82%.

Even with scaled data, both clustering algorithms did not implicitly distinguished observations by hbe weight percentage.


Here are the clusters determined the gibbs sampler on scaled data.

```{r echo=FALSE}
gibb_cols <- fitted_clusters2
# assigning colours to each cluster
gibb_cols[gibb_cols == 1] = "red"
gibb_cols[gibb_cols == 2] = "blue"
gibb_cols[gibb_cols == 3] = "green"
gibb_cols[gibb_cols == 4] = "orange"
gibb_cols[gibb_cols == 5] = "black"
gibb_cols[gibb_cols == 6] = "pink"
pairs(theta_hat,col=gibb_cols, main="pairs plot") # pairs plot
```

As expected, the clusters determined by the gibbs sampler overlap each other heavily.

It is difficult to assess if the gibbs sampler or the km++ fit the data well, because there is no true cluster affiliation to evaluate against. Additionally, the data with the given variables cannot clearly by group up into distinguishable groups.  

Overall, this suggests that given the existing parameters, observations with particular hbe weight percentages are not very distinguishable from other observations with different hbe weight percentages.

This does not rule out the possibility that hbe weight percentages or other variables can be used to group up microparticle trajectories, but for this to occur, latent variables would need be made known.

# Discussion

This paper proposes an alternative to model-based hierarchal clustering that does not utilize any EM algorithms, which become increasingly expensive, but rather opts for a Bayesian approach.

By making some assumptions, we managed to reduce the complex problem of model-base hierarchal clustering into a simpler problem solved by a Gibbs Sampler.

The main benefit to this is that our Gibbs Sampler method is computationally cheaper as we avoid inverting many large matrices yet still performs the same task of hierarchal clustering.

However, as with all Bayesian analysis, our method requires arbitrarily selected prior distributions to begin. This makes any analysis done with this approach subjective.

To showcase our method, we applied it onto the hbe dataset and compared it against k-means plus plus algorithm. The goal of this application is to see if the data can be clustered in such a way that each cluster corresponds to observations with a particular weight percentage. Through the results of both methods, we discovered that observations in the data could not be easily grouped into clusters distinguished by a particular weight percentage variable. This application does not assert the effectiveness of our proposed method or lack thereof.

With that said, we outline one potential direction for research. One particular research direction would be construct similar clustering methods with different gibbs samplers that have different prior distributions and compare the performances against each other. The goal of this research would not be to determine which gibbs sampler method works best, but rather to discover what would be the most appropriate gibbs sampler for particular situations and document them. 



# Appendix

```{r eval = FALSE}
hbe_fit <- readRDS("hbe-fbm_fit_v1.rds")
N <- length(hbe_fit)
p <- length(hbe_fit[[1]]$coef)
# Extract data 
wts <- vector(length=N)
theta_hat <- matrix(nrow=N, ncol=p)
colnames(theta_hat) <- names(hbe_fit[[1]]$coef)
obs_info <- array(dim=c(p, p, N))
for (i in 1:N){
  theta_hat[i, ] <- hbe_fit[[i]]$coef
  wts[i] <- hbe_fit[[i]]$wt
  obs_info[, , i] <- hbe_fit[[i]]$vcov
}
# Transform the wts into numeric labels
lab <- unique(wts) 
K <- length(lab)
cluster <- numeric(length=N)
for (i in 1:N){
  if (wts[i] == lab[1]){
    cluster[i] <- 1
  }
  else if (wts[i] == lab[2]){
    cluster[i] <- 2
  }
  else if (wts[i] == lab[3]){
    cluster[i] <- 3
  }
  else if (wts[i] == lab[4]){
    cluster[i] <- 4
  }
  else if (wts[i] == lab[5]){
    cluster[i] <- 5
  }
  else{
    cluster[i] <- 6
  }
}
# getting true cluster affiliations
real_N_k <- numeric(length=K)
real_cols <- rep(NA,N) # colors for true cluster affiliations
for (k in 1:K){
  real_N_k[k] <- sum(cluster == k)
}
# assigning colours to each cluster
real_cols[cluster == 1] = "red"
real_cols[cluster == 2] = "blue"
real_cols[cluster == 3] = "green"
real_cols[cluster == 4] = "orange"
real_cols[cluster == 5] = "black"
real_cols[cluster == 6] = "pink"
pairs(theta_hat,col=real_cols, main="pairs plot") # pairs plot
```



```{r, fig.width=7, fig.height=5, eval=FALSE}
pairs(theta_hat)
```


```{r results="hide", eval=FALSE}
# fit gibbs sampler
gibbs_fit <- gibbsSampler(data=theta_hat, V=obs_info, burnin_period=1e3, numIter=1e4, K=6, Theta.out=FALSE, z.out=FALSE)
```

```{r, eval=FALSE}
# get estimated cluster affiliation from gibbs sampler
fitted_clusters <- numeric(length=N)
for (i in 1:N){
  fitted_clusters[i] <- which.max(gibbs_fit$Lambda[i, ])
}
# error rate
err <-  sum(cluster != fitted_clusters)/ N
print(paste0("Percentage of estimated cluster memberships that do not match true cluster memberships: ", round(err,5)))
```


```{r, eval=FALSE}
# clustering real data with kmeans++ algorithm
# note since kmeans starts with random centroids, individual kmeans can have different results.
# We will conduct many kmeans and aggregate the results
#' Initial cluster allocation.
#'
#' Initializes the clusters using the kmeans++ algorithm of Arthur & Vassilvitskii (2007).
#'
#' @param y An `N x p` matrix of observations, each of which is a column.
#' @param K Number of clusters (positive integer).
#' @param max_iter The maximum number of steps in the [stats::kmeans()] algorithm.
#' @param nstart The number of random starts in the [stats::kmeans()] algorithm.
#'
#' @return A vector of length `N` consisting of integers between `1` and `K` specifying an initial clustering of the observations.
#' @author Martin Lysy
#' @references Arthur, D., Vassilvitskii, S. "k-means++: the advantages of careful seeding" *Proceedings of the 18th Annual ACM-SIAM Symposium on Discrete Algorithms.* (2007): 1027–1035. <http://ilpubs.stanford.edu:8090/778/1/2006-13.pdf>.
init_z <- function(y, K, max_iter = 10, nstart = 10) {
  # init++
  N <- nrow(y)
  p <- ncol(y)
  x <- t(y) # easier to use columns
  centers <- matrix(NA, p, K) # centers
  icenters <- rep(NA, K-1) # indices of centers
  minD <- rep(Inf, N) # current minimum distance vector
  # initialize
  icenters[1] <- sample(N,1)
  centers[,1] <- x[,icenters[1]]
  for(ii in 1:(K-1)) {
    D <- colSums((x - centers[,ii])^2) # new distance
    minD <- pmin(minD, D)
    icenters[ii+1] <- sample(N, 1, prob = minD)
    centers[,ii+1] <- x[,icenters[ii+1]]
  }
  centers <- t(centers)
  colnames(centers) <- rownames(x)
  # run kmeans with these centers
  km <- kmeans(x = y, centers = centers, nstart = nstart, iter.max = max_iter)
  km$cluster
}
max_iters <- 10000  # set number of kmeans to run
kmeans_results <- matrix(nrow = N,ncol = max_iters) # pre-allocate matrix
# simple function to calculate the mode in vector of numbers
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
# get results from max_iters kmeans
for (iter in 1:max_iters){
  kmeans_results[,iter] <- init_z(y = theta_hat,K = 6)
}
fitted_clusters1 <- matrix(nrow = N,ncol = 1) # pre-allocate matrix to store estimated clusters
# the mode cluster membership for each observation becomes the estimate cluster membership
for (row in 1:N){
  fitted_clusters1[row,] <- Mode(kmeans_results[row,])
}
err1 <- sum(cluster != fitted_clusters1) / N
print(paste0("Percentage of estimated cluster memberships that do not match true cluster memberships: ", round(err1,5)))
```


```{r eval= FALSE}
#-----------------------------------------------------------------------------------------------------------
# testing with standardized data (data centered at 0 with sd of 1)
#-----------------------------------------------------------------------------------------------------------
theta_hat_scaled <- scale(theta_hat)
# fit gibbs sampler
gibbs_fit1 <- gibbsSampler(data=theta_hat_scaled, V=obs_info, burnin_period=1e3, numIter=1e4, K=6, Theta.out=FALSE, z.out=FALSE)
```

```{r eval = FALSE}
# get estimated cluster affiliation from gibbs sampler
fitted_clusters2 <- numeric(length=N)
for (i in 1:N){
  fitted_clusters2[i] <- which.max(gibbs_fit1$Lambda[i, ])
}
# error rate
err2 <- sum(cluster != fitted_clusters2) / N
print(paste0("Percentage of estimated cluster memberships that do not match true cluster memberships: ",round(err2,5)))
```


```{r eval=FALSE}
max_iters <- 10000
kmeans_results1 <- matrix(nrow = N,ncol = max_iters)
# get results from max_iters kmeans
for (iter in 1:max_iters){
  kmeans_results1[,iter] <- init_z(y = theta_hat_scaled,K = 6)
}
fitted_clusters3 <- matrix(nrow = N,ncol = 1) # pre-allocate matrix
# the mode cluster membership for each observation becomes the estimate cluster membership
for (row in 1:N){
  fitted_clusters3[row,] <- Mode(kmeans_results1[row,])
}
err3 <- sum(cluster != fitted_clusters3) / N
print(paste0("Percentage of estimated cluster memberships that do not match true cluster memberships: ",round(err3,5)))
```

```{r eval=FALSE}
gibb_cols <- fitted_clusters2
# assigning colours to each cluster
gibb_cols[gibb_cols == 1] = "red"
gibb_cols[gibb_cols == 2] = "blue"
gibb_cols[gibb_cols == 3] = "green"
gibb_cols[gibb_cols == 4] = "orange"
gibb_cols[gibb_cols == 5] = "black"
gibb_cols[gibb_cols == 6] = "pink"
pairs(theta_hat,col=gibb_cols, main="pairs plot") # pairs plot
```


# References

Charles Bouveyron, Camille Brunet. Model-Based Clustering of High-Dimensional Data: A review. Computational Statistics and Data Analysis, Elsevier, 2013, 71, pp.52-78. ff10.1016/j.csda.2012.12.008ff. ffhal-00750909f

K. M. A. Patel and P. Thakral, "The best clustering algorithms in data mining," 2016 International Conference on Communication and Signal Processing (ICCSP), Melmaruvathur, 2016, pp. 2042-2046.

A. Samé, "Grouped data clustering using a fast mixture-model-based algorithm," 2009 IEEE International Conference on Systems, Man and Cybernetics, San Antonio, TX, 2009, pp. 2276-2280.

R.M. Neal and Hinton, A view of the EM algorithm that justifies incremental, sparse and other variants. In M.I. Jordan, editor, Learning in Graphical Models, 355-368, Kluwer, Dordrecht, 1998

P.S. Bradley, M.U. Fayyad and C.A. Reina, Scaling EM (Expectation Maximization) clustering to large databases. Technical Repport MSRTR-98-35, Microsoft Research, 1998.

B. Zhang, "Comparison of the Performance of Center-Based Clustering Algorithms", Advances in Knowledge Discovery and Data Mining, Lecture Notes in Computer Science, Volume 2637/2003, pp. 569, 2003.

Hill, D.B., Vasquez, P.A., Mellnik, J., McKinley, S.A., Vose, A., Mu, F., Henderson, A.G.,
Donaldson, S.H., Alexis, N.E., Boucher, R.C., and Forest, M.G. (2014).  “A biophysical
basis for mucus solids concentration as a candidate biomarker for airways disease.”
PLoS ONE
, 9(2): e87681

Arthur, D. & Vassilvitskii, S. (2007). k-means++: the advantages of careful seeding.. In N. Bansal, K. Pruhs & C. Stein (eds.), SODA (p./pp. 1027-1035), : SIAM. ISBN: 978-0-898716-24-5

